---
title: "Data Download & Preprocessing Walkthrough"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Download & Preprocessing Walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette walks the full data-prep pipeline (`CLAUDE.md`'s pipeline
order) end to end on a small dummy site, so every function's real inputs
and outputs are visible in one place. See `01-gee-setup` first for GEE/Drive
authentication — every code chunk below assumes that's already done.

The dummy site is a ~1.2 km square near the University of Idaho Arboretum
in Moscow, ID, over two summer months (July-August 2023) — small and
snow-free enough to keep the walkthrough fast if you were to run it for
real, while exercising every step of the pipeline.

```{r}
library(sf)
library(terra)
library(SpencerEcoTools)

study_area <- "UI_Arboretum"

dates.v <- seq(as.Date("2023-07-01"), as.Date("2023-08-01"), by = "month")
#> [1] "2023-07-01" "2023-08-01"
```

## 1. Define the area of interest

```{r}
pt <- sf::st_sfc(sf::st_point(c(-117.0166, 46.7197)), crs = 4326)
pt <- sf::st_sf(id = 1, geometry = pt)

ee_aoi <- define_aoi(
  aoi         = pt,
  buffer_dist = 600,           # meters -> ~1.2 x 1.2 km AOI
  crs_epsg    = "EPSG:32611"   # UTM zone 11N
)
```

`ee_aoi` is a named list (`geometry`, `crs`, `bbox_wgs84`) that every
`download_*()` function below expects as its `aoi` argument.

## 2. Download the DEM

```{r}
download_dem(
  aoi     = ee_aoi,
  out_dir = "./Data/DEM",
  scale   = 30
)

dtm <- terra::rast("./Data/DEM/DEM_GLO30.tif")
```

`download_dem()` doesn't take a `study_area` argument — the output is
always named `DEM_GLO30.tif` in `out_dir`, since a study area typically has
just one DEM shared across every period modeled.

## 3. Download multispectral imagery (HLS or Sentinel-2)

```{r}
download_hls(
  aoi        = ee_aoi,
  dates      = dates.v,
  out_dir    = "./Data/Imagery",
  study_area = study_area
)
```

`download_hls()` blends Harmonized Landsat Sentinel-2 (S30 + L30), which
covers years before 2019. For a study window entirely after ~2019,
`download_s2()` is a simpler Sentinel-2-only alternative with the same
0-1 scaled 4-band output:

```{r}
download_s2(
  aoi        = ee_aoi,
  dates      = dates.v,
  out_dir    = "./Data/Imagery",
  study_area = study_area
)
```

Only one of the two is needed — both write the same `red`/`green`/`blue`/
`nir` band layout that every downstream function below expects.

## 4. Download and downscale LAI

```{r}
download_modis_lai(
  aoi        = ee_aoi,
  dates      = dates.v,
  out_dir    = "./Data/LAI/Coarse",
  study_area = study_area
)

downscale_lai(
  dates      = dates.v,
  img_dir    = "./Data/Imagery",
  lai_dir    = "./Data/LAI/Coarse",
  out_dir    = "./Data/LAI/Fine",
  study_area = study_area
)
```

`downscale_lai()` calls `microclimdata::lai_fromndvi()` internally,
fusing the coarse (500 m) MODIS LAI with NDVI computed from the fine (30 m)
imagery downloaded in step 3 — this is the package's first direct use of
`microclimdata`.

## 5. Download and compute albedo

```{r}
download_albedo(
  aoi        = ee_aoi,
  dates      = dates.v,
  out_dir    = "./Data/Albedo/Modis",
  study_area = study_area
)

compute_albedo(
  dates      = dates.v,
  modis_dir  = "./Data/Albedo/Modis",
  img_dir    = "./Data/Imagery",
  out_dir    = "./Data/Albedo/Fine",
  study_area = study_area
)
```

`compute_albedo()` calls `microclimdata::albedo_fromaerial()` to compute
photographic albedo from the imagery, then `microclimdata::albedo_adjust()`
to rescale it to match the coarse MODIS broadband albedo.

## 6. Land cover, vegetation height, and soil

Land cover and vegetation height each have two sources: this package's own
NLCD/LANDFIRE preprocessing (what you'd already have if you work with
those products), or pulling ready-to-use rasters directly from
`microclimdata`'s own Earth Engine downloads. Both paths feed the same
downstream arguments — pick whichever source you have. Only ESA + ETH is
shown running end-to-end below; the NLCD/LANDFIRE side is included so both
are visible.

### Land cover

```{r}
# (a) Primary: your own NLCD raster, converted to CORINE classes
corine_lc <- NLCD_2_CORINE(
  lc            = terra::rast("./Data/LandCover/NLCD_2021.tif"),
  crop_template = dtm
)

# (b) Alternative: ESA WorldCover pulled directly via microclimdata (no
#     conversion step needed)
esa_lc <- microclimdata::lcover_download(
  r                 = dtm,
  type              = "ESA",
  year              = 2020,
  GoogleDrivefolder = "GEE_Exports",
  pathtopython      = "C:/Users/you/miniforge3/envs/rgee"
)
```

### Vegetation height

```{r}
# (a) Primary: your own LANDFIRE Existing Vegetation Height raster
lf_veg_height <- LandfireVegHght_AsNumeric(
  rast          = terra::rast("./Data/VegHeight/LF2020_EVH.tif"),
  crop_template = dtm
)

# (b) Alternative: ETH Global Canopy Height pulled directly via microclimdata
veg_height <- microclimdata::vegheight_download(
  r                 = dtm,
  GoogleDrivefolder = "GEE_Exports",
  pathtopython      = "C:/Users/you/miniforge3/envs/rgee"
)
```

Whichever land cover source you use determines the `lctype` argument
downstream (`"CORINE"` for NLCD-derived, `"ESA"` for WorldCover) in
`compute_reflectance()` and `package_veg_soil()` below — and the matching
`water` land-cover code (512 for CORINE's water class, 80 for ESA
WorldCover's).

### Soil

```{r}
soil <- download_soil(
  dtm        = dtm,
  landcover  = esa_lc,
  out_dir    = "./Data/Soil",
  study_area = study_area,
  water      = 80    # ESA WorldCover water class code
)
```

`download_soil()` wraps `microclimdata::soildata_download()` (coarse
SoilGrids pull) and `microclimdata::soildata_downscale()` (fine-resolution
downscale using the land cover raster) into one call. Calling those two
`microclimdata` functions directly gives more control — e.g. inspecting or
saving the coarse raster before downscaling:

```{r}
dtm_wgs84 <- terra::project(dtm, "epsg:4326")

soil_coarse <- microclimdata::soildata_download(
  dtm_wgs84,
  pathdir     = "./Data/Soil/tmp",
  deletefiles = FALSE
)
terra::writeRaster(soil_coarse, "./Data/Soil/SoilData_Coarse.tif", overwrite = TRUE)

lc_wgs84  <- terra::project(esa_lc, "epsg:4326")
soil_fine <- microclimdata::soildata_downscale(soil_coarse, lc_wgs84, water = 80)
soil_fine <- terra::project(soil_fine, dtm)
```

## 7. Download AORC climate data and estimate diffuse radiation

```{r}
aorc_log <- download_aorc(
  aoi         = ee_aoi,
  dates       = dates.v,
  out_dir     = "./Data/Weather",
  study_area  = study_area,
  workers     = 2,
  python_path = "C:/Users/you/miniforge3/envs/rgee"
)

estimate_diffuse_rad(
  dates      = dates.v,
  aorc_dir   = "./Data/Weather",
  study_area = study_area,
  workers    = 2
)
```

Both functions keep their output in WGS84 (EPSG:4326) — reprojecting AORC
data *before* `estimate_diffuse_rad()` introduces empty edge cells from
projection warping that corrupt its solar geometry calculations.
Reprojection happens later, inside `package_climate()`.

## 8. Compute leaf and ground reflectance

```{r}
compute_reflectance(
  dates        = dates.v,
  landcover    = esa_lc,
  lai_dir      = "./Data/LAI/Fine",
  alb_dir      = "./Data/Albedo/Fine",
  out_dir_lref = "./Data/Reflectance/Lref",
  out_dir_gref = "./Data/Reflectance/Gref",
  study_area   = study_area,
  lctype       = "ESA"
)
```

`compute_reflectance()` calls `microclimdata::x_calc()` (land-cover-based x
values) and `microclimdata::reflectance_calc()` to solve for leaf (`Lref`)
and ground (`Gref`) reflectance from LAI and albedo.

## 9. (Optional) Multi-year climate normals

If you have multiple years of AORC data and want long-term hourly normals
rather than a specific period's actual weather, `summarize_climate_normals()`
stacks each calendar hour across years:

```{r}
summarize_climate_normals(
  dates      = dates.v,
  aorc_dir   = "./Data/Weather",
  out_dir    = "./Data/ClimateNormals",
  stats      = c("mean", "sd"),
  study_area = study_area
)
```

This dummy example only has one summer of data, so it's skipped below —
`package_climate()` in step 11 uses the actual AORC data directly instead.

## 10. Package vegetation and soil parameters

```{r}
dates_df <- data.frame(
  Start_Dates = as.Date("2023-07-01"),
  End_Dates   = as.Date("2023-08-31")
)

veg_soil_log <- package_veg_soil(
  dates            = dates_df,
  snow_free_months = 7:8,
  landcover        = esa_lc,
  veg_height       = veg_height,
  soil_path        = "./Data/Soil/SoilData_Fine_UI_Arboretum.tif",
  lai_dir          = "./Data/LAI/Fine",
  refl_dir         = "./Data/Reflectance",
  vegpara_dir      = "./Data/VegPara",
  soilpara_dir     = "./Data/SoilPara",
  study_area       = study_area,
  lctype           = "ESA",
  water            = 80
)
```

This produces `./Data/VegPara/UI_Arboretum_VegPara_20230701_to_20230831.RDS`
and the equivalent `SoilPara` file, following the
`{study_area}_VegPara_{period_label}.RDS` naming convention (period labels
are always `YYYYMMDD_to_YYYYMMDD`).

## 11. Package climate data

`package_climate()` needs a template raster to reproject and crop AORC
data to — the vegetation parameter grid just produced is the natural
choice, since it's on the same grid `run_micro_big_nichemap()` expects
downstream:

```{r}
veg_template <- terra::rast(readRDS(
  "./Data/VegPara/UI_Arboretum_VegPara_20230701_to_20230831.RDS"
)$hgt)

package_climate(
  dates        = dates_df,
  aorc_dir     = "./Data/Weather",
  out_dir      = "./Data/Weather_Pkg",
  study_area   = study_area,
  template     = veg_template,
  keep_monthly = TRUE
)
```

This produces `./Data/Weather_Pkg/UI_Arboretum_Climate_20230701_to_20230831.RDS`.

## Next steps

The `02-data-pipeline` outputs — the `VegPara`, `SoilPara`, and `Climate`
RDS files, plus the fine and coarse DEMs — are exactly what
`03-running-microclimf` uses to run the tiled `microclimf` model, locally
or on a SLURM cluster.
