---
title: "Running microclimf Models: Local and Cluster"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running microclimf Models: Local and Cluster}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This picks up where `02-data-pipeline` left off: a packaged climate RDS,
vegetation/soil parameter RDS files, and a fine-resolution DEM for the
UI Arboretum dummy site. It covers tiling, running `microclimf` locally,
and running the same model as a SLURM array job.

## 1. Build the coarse (climate-resolution) DEM

`create_tiles()` and `run_micro_big_nichemap()` need a `coarse_dem` on the
same grid as the packaged climate data — not the fine 30 m DEM. Build it by
resampling the fine DEM onto the climate raster's actual grid:

```{r}
library(terra)
library(SpencerEcoTools)

dem_fine <- terra::rast("./Data/DEM/DEM_GLO30.tif")

clim_rds     <- readRDS("./Data/Weather_Pkg/UI_Arboretum_Climate_20230701_to_20230801.RDS")
clim_template <- terra::unwrap(clim_rds[[1]])

dem_coarse <- terra::resample(dem_fine, clim_template)
terra::writeRaster(dem_coarse, "./Data/DEM/DEM_Coarse_UI_Arboretum.tif", overwrite = TRUE)
```

(Climate RDS files store a wrapped `SpatRaster` — `terra::unwrap()` is
required before use, matching how `package_climate()` writes them.)

## 2. Create tiles

```{r}
dates_df <- data.frame(
  Start_Dates = as.Date("2023-07-01"),
  End_Dates   = as.Date("2023-08-01")
)

tiles <- create_tiles(
  coarse_dem        = dem_coarse,
  fine_dem          = dem_fine,
  dates             = as.Date(c("2023-07-01", "2023-08-01")),
  tile_dims         = c(3, 3),
  snow_modeling     = FALSE,
  return_tiles_rast = TRUE
)
```

`tile_dims` sets the SIZE of each tile in coarse-DEM cells (rows, columns
per tile) — not the number of tiles produced. `create_tiles()` derives the
tile count by dividing the coarse DEM's dimensions by `tile_dims`
(`n_tiles_r <- ceiling(nrow(coarse_dem) / tile_nrow)`, and likewise for
columns). With the default `buffer_size = 1`, the smallest valid tile size
is `c(3, 3)` — `create_tiles()` requires `tile_nrow`/`tile_ncol >=
2 * buffer_size + 1`, so `c(1, 1)` errors out immediately. `c(3, 3)`
happens to be large enough to cover this dummy site's roughly 2 x 2
coarse (climate-resolution) grid in a single tile
(`ceiling(2 / 3) = 1` in each dimension). On a real, larger study area,
either pass a bigger `tile_dims` or omit it entirely (`NULL`, the default)
to let `create_tiles()` pick the largest tile size that fits within
`mem_fraction` (default 70%) of available system RAM — this is what keeps
`run_micro_big_nichemap()` memory-safe on large domains.

## 3. Run locally

```{r}
log_local <- run_micro_big_nichemap(
  tiles       = tiles,
  clim        = "./Data/Weather_Pkg",
  dates       = dates_df,
  dtm_fine    = dem_fine,
  dtm_coarse  = dem_coarse,
  vegp        = "./Data/VegPara",
  soilc       = "./Data/SoilPara",
  output_dir  = "./Microclim_out",
  reqhgt      = 2,
  zref        = 2,
  windhgt     = 10,
  matemp      = NA,
  snow        = FALSE,
  Dynreqhgt   = TRUE,
  altcorrect  = 2,
  parallel    = TRUE,
  ncores      = 4,
  study_area  = "UI_Arboretum",
  file_fmt    = "nc",
  compression = 6
)
```

This models every height in one process: the above-ground height
(`reqhgt = 2`, i.e. 2 m) plus all ten fixed soil depths (0-200 cm), for
every tile and every period in `dates`. With no `clust_array_arg`/
`clust_array_size` supplied, all `(tile, period)` combinations run in this
one R session.

## 4. Run on a SLURM cluster

`run_micro_big_nichemap()` never submits jobs itself — `clust_array_arg`
and `clust_array_size` are hidden `...` arguments an SBATCH array script
passes in, so each array task processes a slice of the same
`(tile, period)` combinations the local run processed all at once. Put the
model call in its own script:

`run_microclimf_cluster.R`:
```r
#!/usr/bin/env Rscript
library(terra)
library(SpencerEcoTools)

n_threads <- as.integer(Sys.getenv("SLURM_CPUS_PER_TASK", unset = "1"))
terraOptions(threads = n_threads, memfrac = 0.7, tempdir = "~/terra_temp")

args <- commandArgs(trailingOnly = TRUE)
get_arg <- function(name, default = NA) {
  x <- grep(paste0("^--", name, "="), args, value = TRUE)
  if (length(x) == 0) return(default)
  sub(paste0("^--", name, "="), "", x)
}
clust_array_arg  <- get_arg("clust_array_arg")
clust_array_size <- get_arg("clust_array_size")
if (is.na(clust_array_arg) || is.na(clust_array_size)) {
  stop("Both --clust_array_arg and --clust_array_size are required (missing or ",
       "misspelled flag?). Got clust_array_arg=", clust_array_arg,
       ", clust_array_size=", clust_array_size)
}

dem_fine   <- rast("./Data/DEM/DEM_GLO30.tif")
dem_coarse <- rast("./Data/DEM/DEM_Coarse_UI_Arboretum.tif")

dates_df <- data.frame(
  Start_Dates = as.Date("2023-07-01"),
  End_Dates   = as.Date("2023-08-01")
)

tiles <- create_tiles(
  coarse_dem = dem_coarse, fine_dem = dem_fine,
  dates      = as.Date(c("2023-07-01", "2023-08-01")),
  tile_dims  = c(3, 3), return_tiles_rast = TRUE
)

log <- run_micro_big_nichemap(
  tiles = tiles, clim = "./Data/Weather_Pkg", dates = dates_df,
  dtm_fine = dem_fine, dtm_coarse = dem_coarse,
  vegp = "./Data/VegPara", soilc = "./Data/SoilPara",
  output_dir = "./Microclim_out",
  reqhgt = 2, zref = 2, windhgt = 10, matemp = NA,
  snow = FALSE, Dynreqhgt = TRUE, altcorrect = 2,
  parallel = TRUE, ncores = n_threads,
  study_area = "UI_Arboretum", file_fmt = "nc", compression = 6,
  clust_array_arg  = as.integer(clust_array_arg),
  clust_array_size = as.integer(clust_array_size)
)

cat(sprintf("Task %s/%s complete.\n", clust_array_arg, clust_array_size))
```

And a companion SBATCH script that drives it as an array job. Suppose a
larger real study area produces 4 tiles over 1 period — 4 total
`(tile, period)` tasks, so the array is sized to 4 (the dummy site above,
with its `c(3, 3)` `tile_dims`, would still produce a single tile):

`run_microclimf.sbatch`:
```bash
#!/bin/bash
#SBATCH --job-name=microclimf_UI_Arboretum
#SBATCH --array=1-4
#SBATCH --cpus-per-task=4
#SBATCH --mem=16G
#SBATCH --time=02:00:00
#SBATCH --output=logs/microclimf_%A_%a.out

module load R

Rscript run_microclimf_cluster.R \
  --clust_array_arg=${SLURM_ARRAY_TASK_ID} \
  --clust_array_size=4
```

`run_micro_big_nichemap()` enumerates all `(tile, period)` combinations
1-N and distributes them round-robin across the array
(`rep(seq_len(clust_array_size), length.out = N_tasks)`, tile index
cycling fastest) — so `--array=1-4` here means each of the 4 array tasks
gets exactly one `(tile, period)` combination in this example. Size
`--array` and `--clust_array_size` to `N_tiles x N_periods`, or fewer if
running multiple combinations per node.

## 5. Stitch tile outputs back together

Whether the run was local or a completed SLURM array, each `(tile, period)`
combination is written as a separate file under `output_dir`.
`stitch_tiles_runmicro()` assembles them into one full-domain file per
period and data type — internally it calls the lower-level `stitch_tiles()`
once per period/data-type combination, which is also useful directly if
you want to stitch a single tile directory (rather than a full
`run_micro_big_nichemap()` output tree):

```{r}
stitch_tiles_runmicro(
  output_dir = "./Microclim_out",
  dates      = dates_df,
  study_area = "UI_Arboretum",
  file_fmt   = "nc"
)
```

With `file_fmt = "nc"`, this builds a GDAL VRT per variable (no data
copied, tile files must stay at their original paths); `file_fmt = "h5"`
instead builds an HDF5 virtual dataset. Use whichever `file_fmt` matches
what `run_micro_big_nichemap()` was run with.

## 6. Inspect results at a point

`micro_to_csv()` extracts one grid cell's modeled time series from a
stitched above-ground + below-ground output pair, reshaped into
NicheMapR-style data frames — useful for a quick sanity check, or as input
to point-model workflows downstream of this package's scope:

```{r}
pt_series <- micro_to_csv(
  abvgrd_input    = "./Microclim_out/UI_Arboretum/Stitched/20230701_to_20230801/UI_Arboretum_AbvGrd_MicroclimModel_20230701_to_20230801",
  blwgrd_input    = "./Microclim_out/UI_Arboretum/Stitched/20230701_to_20230801/UI_Arboretum_BlwGrd_MicroclimModel_20230701_to_20230801",
  cell            = 1,
  cell_input_type = "cellnumber",
  dates           = as.Date(c("2023-07-01", "2023-07-07")),
  elev            = 790
)
```

Note the `.vrt`-mode inputs above are the shared *stem* path — no
extension, no variable suffix — since `stitch_tiles_runmicro(file_fmt =
"nc")` writes one `{stem}_{varname}.vrt` file per variable rather than one
combined file.

## Next steps

That completes the pipeline arc across all three vignettes: `01-gee-setup`
authenticated GEE and Drive, `02-data-pipeline` turned raw GEE/AORC/soil
sources into packaged climate, vegetation, and soil parameter grids, and
this vignette tiled those grids, ran the `microclimf`/NicheMapR model
locally or across a SLURM array, stitched the per-tile outputs back into
full-domain files, and extracted an inspectable point time series from the
result. From here, `pt_series` (or the stitched `.nc`/`.h5` files directly)
is ready for whatever downstream analysis — habitat modeling, energetics,
NicheMapR point workflows — motivated the run.
