---
title: "Writing Endotherm Model Inputs: endo.dat, alomvars.dat, JULDAYS.dat"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Writing Endotherm Model Inputs: endo.dat, alomvars.dat, JULDAYS.dat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This is the first step in the endotherm modeling workflow: building the
fixed-format input files (`endo.dat`, `alomvars.dat`, `JULDAYS.DAT`) that the
NicheMapR Endotherm model executable (`Endo2022a.exe`) reads. Everything on
this page runs in plain R and does not require the executable itself — it
picks up in `vignette("05-metabolic-chamber")`.

## 1. Start from the package's reference defaults

`get_endotherm_defaults()` returns the same nine argument groups that
`write_endotherm_inputs()` takes (`model_settings`, `animal`, `fur`,
`physiology`, `diet`, `thermoreg`, `flying_digging`, `nest_shelter`,
`allometry`), pre-filled with a reference Female Bighorn Sheep
parameterization extracted from a real NicheMapR Endotherm run.

```{r}
library(SpencerEcoTools)

defaults <- get_endotherm_defaults()
str(defaults, max.level = 1)
```

## 2. Customize for a study animal

Edit whichever fields differ for your animal. Any field that varies by
Julian day (suffixed `2`, e.g. `mass2`) must stay the same length as
`model_settings$julnum` (12 by default, one value per month).

```{r}
defaults$animal$species <- "Male Bighorn"
defaults$animal$mass    <- 60
defaults$animal$mass2   <- rep(60, defaults$model_settings$julnum)
```

## 3. Write endo.dat and alomvars.dat

`write_endotherm_inputs()` takes the nine argument groups directly, so the
edited `defaults` list can be spread into it with `do.call()`.

```{r}
output_dir <- tempfile("endotherm_inputs_")
dir.create(output_dir)

write_log <- do.call(write_endotherm_inputs, c(list(output_dir = output_dir), defaults))
write_log
```

```{r}
list.files(output_dir)
```

Both files are fixed-format text, parsed by column position — they are not
meant to be hand-edited, but the first few lines of `endo.dat` show the
general shape:

```{r}
head(readLines(file.path(output_dir, "endo.dat")), 5)
```

## 4. Write JULDAYS.dat

`write_juldays_dat()` shares `model_settings` with `write_endotherm_inputs()`
(specifically `julnum`/`juldays`) so the day list always matches what was
just written into `endo.dat` — a mismatch between the two causes the exe's
fixed-format reader to misparse the file. It takes its own
`habitat_settings` group (substrate absorptivity, percent shade, percent
surface wet); the defaults below describe open, mostly unshaded terrain.

```{r}
juldays_log <- write_juldays_dat(
  output_dir,
  model_settings = defaults$model_settings
)
juldays_log

readLines(file.path(output_dir, "JULDAYS.DAT"))
```

## Next steps

`endo.dat`, `alomvars.dat`, and `JULDAYS.DAT` are everything
`Endo2022a.exe` needs for a standard run. `vignette("05-metabolic-chamber")`
picks up from here, running a metabolic chamber calibration against this
same animal parameterization.

```{r, include = FALSE}
unlink(output_dir, recursive = TRUE)
```
